Communicative reduction in referring expressions in a multi-player negotation game

### 2023-01-18 write-ups/cogsci23_paper/manuscript.Rmd
apply Mike's advice

Successful communication is grounded in a shared understanding of what utterances mean in the context they are produced in. In many cases conventional word meanings are enough, but there are also contexts where objects to be referred to are not easy to distinguish. In these situations, establishing reference is more difficult, but no less important.

The formation of these ad-hoc referring expressions has been studied extensively in the context of dyadic reference games [@clarkReferringCollaborativeProcess1986, ADD MORE CITES]. In these, two participants see a set of images (often abstract shapes) and one person is tasked with identifying an image so the other person can pick it out. People can be very successful at this task, achieving high levels of matches. Over repeated reference to the same images, pairs develop shorthand names for the images, leading to shorter descriptions in later repetitions. These nicknames conventionalize with a pair, as they tend to stick to the same description, but are idiosyncratic and differ between pairs.

This has been a well-studied microcosm for understanding reference which has proven useful for testing theories about how referring expressions originate, how expressions are designed, etc [TODO CITES]. The implicit theoretical claim is that these phenomena hold whenever people interact repeatedly in ways that require reference to some objects that do not have pre-conventionalized names that are adequate to distinguish them in context.

However, there are some ways in which this microcosm is not representative of typical language use. For one, in everyday communication, establishing reference rarely happens in isolation but is usually a subgoal in a larger conversational goal. The overall goal might be cooperative, such as in asking for a piece of kitchen equipment while cooking together, but it can also be adversarial, such as in a negotiation where the assets to be divied up must be referenced. In the reference game context, the goal is always cooperative, to match the images or get them in the same order.

TODO paragraph on on the issue of shared vs. privileged knowledge, which has been studied extensively in communication psycholinguistics (e.g., heller, keysar, barr) but IIRC not in the context of creating novel referring expressions in the tangram task. so we are blending in this other important element of naturalism, albeit only around the values and not the actual stuff?

@@

@mankewitz2021 tried to bridge the gap between datasets of negotiation over easily nameable objects, and the reference game literature where there is reference (but no negotiation) over hard to name objects. They created a 3 person game where players each select what flowers to grow from a set of 6 flowers with variable and partially hidden value. Players got the value of the flower only if they were the only one to select it, which incentivized coordination and negotiation between players. However, all the flowers shown were the same color, so they were not easily nameable, and the flowers repeated, leading to opportunities for ad-hoc conventionalization. This game structure encouraged players to describe flowers while never explicitly saying that a player should refer to any specific flower. They also had two conditions, one where players within a group won points together (and thus had fully aligned incentives) and another where they won points as individuals (and didn't). @mankewitz2021 found a slight decrease in language over the course of the game, but they did not conduct a detailed analysis to assess whether the decreases in words were associated with shorter and more conventionalized referring expressions.

Historically, studies of convention formation have had to rely on proxies such as reduction and fluency, qualitatively assess similarity of utterances, or manually measure lexical overlap in small datasets to get at the ideas of convergence and divergence. More recent work has taken advantage of natural language processing tools for mapping words and sentences to semantic vector spaces in order to quantify semantic similarities between sentences [@hawkins2020, @boyce2022]. We follow this recent work in using vector embeddings to measure semantic distances between referring expressions in the @mankewitz2021 dataset.

## Current work
Here we ask whether participants in a complex, multi-player strategic game with differing incentives and shared knoweldge were able to develop conventionalized reference expressions. We explored whether the phenomena found in classic dyadic iterated reference games extended to this more naturalistic domain, and whether they differed across the individual and shared incentive conditions. The key questions we addressed were:

1. Do the referring expressions reduce over time?

2. How do referring expressions change within and between groups over the course of the games?

3. How do the different incentive conditions influence the referring expressions?

@@

### Comparison with post-game names

@@

We also examined how referring expressions related the conventionalized names for other flowers. Utterances did not significantly diverge from the conventionalized names of other flower over time (trial: `r stats_text(to_e_bet_f,6)`, see Figure \ref{toend} lower panels).  There was not an effect of being said by the same person who gave the convention (`r stats_text(to_e_bet_f,8)`), or interactions between who said it and trial number (`r stats_text(to_e_bet_f,7)`).

@@

First, we saw reduction, as referring expressions tended to decrease in length over the course of the game. Second, we found the expected trifecta of convergence and divergence results. Over the course of games, references to the same flower within a game converged both to other descriptions used at that timepoint and to the post-game descriptions. This convergence was specific to flower and games, as references to different flowers within a game and  references to the same flower across games both diverged.

These results align with the expected pattern of partner-specific convention formation established in dyadic reference games. These patterns of reduction and convergence confirm some generalizability of the reference game patterns to the freer form and more naturalistic domain of a negotiation game. While still an artificial situation, this game differs from classic reference games in that the set-up was symmetric, with all players having equivalent amounts of knowledge and authority, and that the need for reference was embedding within a more complex goal structure. This game also was aytpical for reference games in that it had more than 2 players [although see @yoonAudienceDesignMultiparty2019;@boyce2022] and used photos as target images [although see @weber2003].

Our last consideration was how these reduction and convergence effects might be moderated by the shared and individual utility conditions. We found slight condition differences where being in the shared utility condition led to stronger alignment between players throughout. We hypothesize that they might be due to a perceived greater importance of communicating with each other or paying greater attention to what groupmates are seeing. In addition, groups in the shared utility condition diverged less from each other. This reduced divergence suggests that their descriptions and conventions were less ideosyncratic and perhaps closer to shared underlying priors about how to describe the images. We hypothesize that the explicitly more cooperative goal might have led to participants caring more about being understood by their partners. This is interesting in light of @guilbeault2021 finding that larger groups tend to all converge to similar conventions whereas smaller groups can support ideosyncratic conventions.

@@

In conclusion, this analysis showed that the phenomenon of reduction and conventionalization occurs even in complex games with different incentives, asymmetric knowledge, and more open-ended negotiation and dialogue than that found in reference games. Additionally, we saw hints that different incentives may lightly alter moderate the formation of conventions, but that reduction and convention formation took place regardless of incentive condition.

>>>>

Successful communication is grounded in a shared understanding of what utterances mean in the context they are produced in. Whether the communicative goals are cooperative or adversarial, interlocuters need to establish joint reference to be effective. In many cases conventional word meanings are enough, but there are also contexts where objects to be referred to are not easy to distinguish. In these situations, establishing reference is more difficult, but no less important. <!--For one, in everyday communication, establishing reference rarely happens in isolation but is usually a subgoal in a larger conversational goal. The overall goal might be cooperative, such as in asking for a piece of kitchen equipment while cooking together, but it can also be adversarial, such as in a negotiation where the assets to be divied up must be referenced. In the reference game context, the goal is always cooperative, to match the images or get them in the same order. -->

The formation of these ad-hoc referring expressions has been studied extensively in the context of dyadic reference games [@clarkReferringCollaborativeProcess1986, ADD MORE CITES]. In these, two participants see a set of images (often abstract shapes) and one person is tasked with identifying an image so the other person can pick it out. People can be very successful at this task, achieving high levels of matches. Over repeated reference to the same images, pairs develop shorthand names for the images, leading to shorter descriptions in later repetitions. These nicknames conventionalize with a pair, as they tend to stick to the same description, but are idiosyncratic and differ between pairs. This has been a well-studied microcosm for understanding reference which has proven useful for testing theories about how referring expressions originate, how expressions are designed, etc [TODO CITES].

The implicit theoretical claim is that reduction to partner-specific conventions occurs whenever people interact repeatedly in ways that require reference to some objects that do not have adequate pre-conventionalized names. Do these phenomena generalize?

Our current work begins to address this by looking for these phenomena in a slightly more naturalistic setting. One way in which reference games have not been representative of typical language use is in there isolated goal of reference. In everyday communication, establishing reference rarely happens in isolation, but is often an instrumental subgoal in a larger conversation. The overall task may be cooperative, such as asking for kitchen equipment while cooking together, or adversarial, or such identifying items while negotiating a division of assets. In the reference game context, the goal is explicitly cooperative: match corresponding images.

@@

@mankewitz2021 started to bridge the gap between reference games with hard-to-name objects and negotation datsets wtih easily nameable objects. They created a 3 person game where players each select what flowers to grow from a set of 6 flowers with variable and partially hidden value. Players got the value of the flower only if they were the only one to select it, which incentivized coordination and negotiation between players. However, all the flowers shown were the same color, so they were not easily nameable, and the flowers repeated, leading to opportunities for ad-hoc conventionalization. This game structure encouraged players to describe flowers while never explicitly saying that a player should refer to any specific flower. They also had two conditions, one where players within a group won points together (and thus had fully aligned incentives) and another where they won points as individuals (and didn't). @mankewitz2021 found a slight decrease in language over the course of the game, but they did not conduct a detailed analysis to assess whether the decreases in words were associated with shorter and more conventionalized referring expressions. Here, we separate out the referring expressions from other language produced in order to check for patterns of convention-formation.

Our current work addresses whether participants in a complex, multi-player strategic game with differing incentives and shared knowledge were able to develop conventionalized reference expressions. We explored whether the phenomena found in classic dyadic iterated reference games extended to this more naturalistic domain, and whether they differed across the individual and shared incentive conditions. The key questions we addressed were:

1. Did the referring expressions reduce over time?

2. How did referring expressions change within and between groups over the course of the games?

3. How did the different incentive conditions influence the referring expressions?

Historically, studies of convention formation have had to rely on proxies such as reduction, subjective judgements of similarity, and and manual measures of lexical overlap. In contrast, we use natural language processing tools for mapping sentences to semantic vector spaces to quantify semantic similarities between sentences, following recent work by @hawkins2020 and @boyce2022.

@@

### Comparison with post-game convention reports

@@

We also examined how referring expressions related the conventionalized names for other flowers. Utterances diverge from the conventionalized names of other flowers over time (trial: `r stats_text(to_e_bet_f,6)`, see Figure \ref{toend} lower panels).  There was not an effect of being said by the same person who gave the convention (`r stats_text(to_e_bet_f,8)`), or interactions between who said it and trial number (`r stats_text(to_e_bet_f,7)`).

@@

First, we saw reduction, as referring expressions tended to decrease in length over the course of the game. Second, we found the expected set of convergence and divergence patterns. References to the same flower within a game converged both to other descriptions used at that timepoint and to the post-game descriptions. This convergence was specific to flower and games, as references to different flowers within a game and references to the same flower across games both diverged.

These patterns of reduction and convergence confirm some generalizability of the reference game patterns to the freer form and more naturalistic domain of a negotiation game. While still an artificial situation, this game differed from classic reference games in that the set-up was symmetric, with all players having equivalent amounts of knowledge and authority, and that the need for reference was embedding within a more complex goal structure. This game also was aytpical for reference games in that it had more than 2 players [although see @boyce2022; @yoonAudienceDesignMultiparty2019] and used photos as target images [although see @weber2003].

Our third consideration was how the different utility conditions moderated the reduction and convergence effects. We found slight condition differences where being in the shared utility condition meant more similar descriptions for the same flowers and more different descriptions for different. In addition, groups in the shared utility condition diverged less from each other. This pattern of condition differences is consistent with less idiosyncratic descriptions in the shared utility condition, possibly due to a greater concern with being understood by groupmates. The greater alignment between players (and between games) might be due to a perceived greater importance of communicating with each other and understanding what others say, possibly triggered by the cooperative goal. The similarity across games is particularly intriguing given @guilbeault2021, which found  that larger groups tend to all converge to similar conventions whereas smaller groups can support idiosyncratic conventions.

@@

This dataset is just the start of testing how well findings from reference games generalize to more complex conversational situations. One dimension that we were unable to explore here due to the design of the game is the idea of shared versus privileged knowledge (CITATIONS). In the flowers game, like in reference games, all players saw the same set of images (and knew they did), which makes making descriptions to flowers easier. In real-world situations, interlocuters may have different sets of objects in their context and may not know what is in their interlocuters context. An important future direction is studying how repeated reference phenomena interact with different communication forms such as goals and contexts.

In conclusion, this analysis showed that the phenomenon of reduction and conventionalization occurs even in complex games with different incentives, asymmetric knowledge, and more open-ended negotiation and dialogue than that found in reference games. Additionally, we saw hints that different incentives may lightly alter moderate the formation of conventions, but that reduction and convention formation took place regardless of incentive condition. We hope future work will continue to develop our understanding of these phenomena in more natural settings.
